---
title: 'Tutorial 02'
author: 'André Victor Ribeiro Amaral'
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

<!-- > The solution can be found [here](./sol_NNefvA.html). -->

## Introduction {-}

For this problem set, we will use the following libraries

```{r, results='hide', message=FALSE, warning=FALSE}
library(tidyverse)
library(lubridate)
library(zoo)
library(rgeoboundaries)
library(patchwork)
library(viridis)
```

Throughout the tutorial, consider the following scenario. Let $\{y_{it}; i = 1, \cdots, I; t = 1, \cdots, T\}$ denotes a multivariate time series of disease counts in each Saudi Arabia region, such that $I = 13$ refers to the number of considered regions and $T = 36$ to the length of the time series. In particular, we will model $Y_{it}|Y_{i,(t-1)} = y_{i,(t-1)} \sim \text{Poisson}(\lambda_{it})$, such that $\log(\lambda_{it}) = \beta_0 + e_{i} + \beta_1 \cdot t + \beta_2 \cdot \sin(\omega t) + \beta_3 \cdot \cos(\omega t)$, where $e_i$ corresponds to the population fraction in region $i$ and $\omega = 2\pi / 12$. Also, we have information about the population size and proportion of men in each region, as well as the number of deaths linked to each element of the $\{y_{it}\}_{it}$ series.

You can download the `.csv` file [here](./others/r4ds.csv). And you read the file and convert it to a `tibble` object in the following way

```{r}
data <- readr::read_csv(file = 'others/r4ds.csv')
head(data, 5)
```

The code to generated such a data set can be found [here](./others/codeR.R).

$\phantom{-}$

## Problem 01 {-}

Now, to manipulate our data set, we will use the `dplyr` package. From [its documentation](https://dplyr.tidyverse.org/), we can see that there are 5 main methods, namely, `mutate()`, `select()`, `filter()`, `summarise()`, and `arrange()`. We will use them all.

- Start by selecting the `date`, `region`, and `n_cases` from our original data set. To do this, use the pipe operator `%>%` (in `RStudio`, you may use the shortcut `Ctrl + Shift + M`).

- Using `?tidyselect::select_helpers`, one may find useful functions that can be combined with `select()`. For instance, select all columns that start with "`n_`".

- Aiming to obtain more meaningful sliced data sets, use the `filter()` and `select()` functions to select the the dates and regions for which the number of cases is greater than 40.

- Also, select the date, name of the region, number of cases, and number of deaths for which the region is Eastern or Najran `AND` the date is greater than `2019-01-01` (you may want to use the `lubridate` package to deal with dates). 

- Now, select the date, name of the region, number of cases, and number of deaths for which the number of cases is larger than `700` `OR` the number of deaths is equal to `10` and arrange, in a descending order for `n_cases`, the results.

- Next, using the `mutate()` function, create a new column into the original data set that shows the cumulative sum for the number of deaths (and name it `cum_deaths`), and select the number of deaths and this newly created column.

- Using the `lag()` and `drop_na()` functions, create a new column named `n_cases_lag2` and another one named `n_deaths_lag3` that are copies of `n_cases` and `n_deaths` but with lags 2 and 3, respectively. Then, drop the rows with `NA`s in the `n_cases_lag2` column and select all but the `date` and `region` columns.

- Now, the goal is work with the `group_by()` function. To do this, group the data set by region, and select the three first columns.

- Combining a few different functions, select all but the `pop` column and create a new one (named `norm_cases`) that shows the normalized number of cases by date.

- Using the `summarise()` (or `summarize()`) function, get the average and variance for the variables `n_cases` and `n_deaths`, as well as the total number of rows.

- You will notice that the results from the above item do not seem very impressive. However, we can combine `summarize()` with `group_by()` to get more meaningful statistics. For instance, obtain similar results as before, but for each region.

- And as in `SQL`, in `tidyverse`, there are also functions that make joining two tibbles possible. From the [documentation page](https://dplyr.tidyverse.org/reference/mutate-joins.html), we have `inner_join()`, `left_join`, `right_join()` and `full_join()`. As an example, consider the following data set, which shows the (fake) mean temperature (in Celsius) in almost all regions and studied months in Saudi Arabia. Assuming the previous data set is in `data`, 

```{r}
set.seed(0)
n <- (nrow(data) - 3)
new_data <- tibble(id = 1:n, date = data$date[1:n], region = data$region[1:n], temperature = round(rnorm(n, 30, 2), 1))
new_data
```

Notice that there is a column named `id` on `new_data`. We will need it to link the new tibble to the old one. Thus, let's create a similar column in `data`.

```{r}
data <- data %>% add_column(id = 1:(nrow(data))) %>% select(c(7, 1:6))
head(data, 1)
```

Finally, incorporate the new data into the original tibble by using, for example, the `left_join()` function.

$\phantom{-}$

## Problem 02 {-}

For this problem, we will use `ggplot2` to produce different plots. 

- But before, from the original data set, create an object that contains the columns `date`, `region`, `n_men`, `n_women`, `n_cases`, and `n_deaths`, such that `n_men` and `n_women` correspond to the number of men and the number of women, respectively (to do this, recall that `men_prop` represents the proportion of men in the population for a given region).

- Now, the main goal is visualize the studied data set. To do this, we will mainly rely on the `ggplot2` library, which is also included in the `tidyverse` universe. Start by creating a line chart for the number of cases in each region (all in the same plot) over the months (when I tried to plot it in my `RMarkdown` document, I had problems with the `Ha’il` name. If the same happens to you, try to replace it by `Hail`). Map `region` to `color`.

- When plotting, we can also map more than one property at a time. For instance, plot a dot chart for the number of cases in each region over months that maps `region` to `color` and `n_deaths` to `size`.

- Now, using the function `rollmean()` from the `zoo` package, plot the number of cases in Mecca (notice that you will have to filter your data set) over time and the moving average (with window $\text{k} = 3$) of the same time series.

---

Now, we will plot the map of Saudi Arabia divided into the given regions. To do this, we will need to work with a `shapefile` for the boundaries of the Kingdom. This type of file may be found online. However, there are some `R` packages that provide such data. For instance, we can use the `geoboundaries('Saudi Arabia', adm_lvl = 'adm1')` function from the `rgeoboundaries` package to obtain the desired data set.

After downloading the data, you can plot it using the `ggplot(data) + geom_sf()` functions (depending on what you are trying to do, you may have to install (and load) the `sp` and/or `sf` libraries).

```{r, warning=FALSE}
KSA_shape <- geoboundaries('Saudi Arabia', adm_lvl = 'adm1')
KSA_shape$shapeName <- data$region[1:13]
ggplot(data = KSA_shape) + 
  geom_sf() + 
  geom_sf_label(aes(label = shapeName)) +
  theme_bw() + 
  labs(x = 'Longitude', y = 'Latitude', title = 'Map of Saudi Arabia') + 
  theme(text = element_text(family = 'LM Roman 10'))
```

- As a final task, plot two versions of the above map side-by-side. On the left, the colors should represent the population size for men in each region; similarly, on the right, the population size for women (to combine plots in this way, you may want to use the `patchwork` package. And for different color palettes, you may use the `viridis` package). Use the same scale, so that the values are comparable.
